Splitting It Up: The \pkg{spduration} Split-Population Duration Regression Package

### 2016-03-11 manuscript2.tex
added discussion and cites for other packages

One interesting extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \pkg{survival} to the \pkg{spduration} library. This would require some modification of the class as our model also requires information on the risk status of spells, but would give access to the much broader functionality, especially for descriptive summaries of data, in the \pkg{survival} library.

Otherwise, the \pkg{smcure} package \citep{cai2012smcureR}, described in \citet{cai2012smcure}, appears to be the only alternative split-population/cure rate duration model implementation in \proglang{R}. It implements semi-parametric proportional hazards and accelerated failure time cure models using methods outlined in \citep{peng2003fitting} and \citep{zhang2007new}. The semi-parametric estimation of the hazard function shape requires no assumption about its shape, which can be difficult in practice to empirically evaluate, as the authors of \pkg{smcure} discuss. However, \pkg{spduration} offers two advantages over \pkg{smcure}. First, it is much less limited in practical application as it allows for the inclusion of time-varying covariates. Second, it features more post-estimation methods, including those discussed above to create and evaluate model predictions.

Finally, \proglang{Stata} implements cure models similar to our own, and which can handle time-varying covariates, but this is of course of limited appeal to \proglang{R} users.

>>>>

One useful extension of our project in the future would be to integrate and adapt the \code{Surv} class for survival data in \pkg{survival} to the \pkg{spduration} library. This would require some modification of the class as our model also requires information on the risk status of spells, but would give access to the much broader functionality, especially for descriptive summaries of data, in the \pkg{survival} library.

Other currently available \proglang{R} routines for estimation of split-population/cure rate duration models include the packages \pkg{smcure}  \citep{cai2012smcureR}, described in \citet{cai2012smcure}, and \pkg{nltm} \citep{garibotti2010nltm}. \pkg{smcure} implements semi-parametric proportional hazards and accelerated failure time cure models using estimation procedures presented in \citet{peng2003fitting} and \citet{zhang2007new}.\footnote{There is also the older \pkg{gfcure} package, which estimates the parametric accelerated failure time cure model discussed in \citet{peng1998generalized}. A version for Windows is available at \url{http://post.queensu.ca/~pengp/software.html}.} The package \pkg{nltm} will estimate a semi-parametric proportional hazard cure model, and a number of other survival models, using the approach developed in \citet{tsodikov2003semiparametric,tsodikov2007profile}. Semi-parametric estimation of the hazard function is attractive because it requires no assumption about its shape, and such assumptions can be difficult in practice to empirically evaluate. The main advantage \pkg{spduration} offers is the ability to include time varying covariates, which gives it a broader range of practical applications, as \pkg{smcure} and\pkg{nltm} will only accommodate case-based data. \pkg{spduration} also features more post-estimation methods than alternative packages, including those discussed above to create and evaluate model predictions.

Finally, there are \proglang{Stata} routines for implementing cure models \citep[see, e.g.,][]{lambert2007modeling,buxton2013cureregr}, some of which can handle time-varying covariates, but this is of course of limited appeal to \proglang{R} users.
